---
title: "Project"
author: "Het Patel"
date: "2022-10-12"
output: pdf_document
fontsize: 12pt
bibliography: "references.bib"
params:
  data: project2022.csv
---

```{r setup, include=FALSE}

library(tidyverse)
library(ggpubr)
library(janitor)
library(knitr)

```

```{r data 1, include=FALSE}
project <- read.csv(params$data)

a <- project %>%
  select("height", "weight") #this filters the dataset to only have the 'height' and 'weight' columns

b <- project %>%
  select("gender", "height") #this filters the dataset to only have the 'height' and 'weight' columns

test2_male <- b %>%
  filter(gender == "Male")

test2_female <- b %>%
  filter(gender == "Female")

c <- project %>%
  select("gender", "phys") #this filters the dataset to only have the 'height' and 'weight' columns

female_none <- length(which(c$gender=="Female"&c$phys=="None"))
female_none
female_moderate <- length(which(c$gender=="Female"&c$phys=="Moderate"))
female_moderate
female_intense <- length(which(c$gender=="Female"&c$phys=="Intense"))
female_intense

male_none <- length(which(c$gender=="Male"&c$phys=="None"))
male_none
male_moderate <- length(which(c$gender=="Male"&c$phys=="Moderate"))
male_moderate
male_intense <- length(which(c$gender=="Male"&c$phys=="Intense"))
male_intense

gender_phys <- matrix(c(female_none, male_none, female_moderate, male_moderate, female_intense,male_intense), ncol=2,byrow=TRUE)
colnames(gender_phys) <- c("Female", "Male")
rownames(gender_phys) <- c("None","Moderate","Intense")
gender_phys <- as.table(gender_phys)
```

# Abstract:

## Linear Regression Test:

This test was conducted in order to determine whether there was a linear
relationship between height and weight.  
\
**Findings:**

|                        | Test Statistic | Degrees of Freedom | P-Value             | 95% Confidence Interval |
|------------------------|----------------|--------------------|---------------------|-------------------------|
| Linear Regression Test | 17.3341348     | 998                | 4.8071712 x 10\^-59 | (0.2648081, 0.3324183)  |

Therefore, it is determined that there is a linear relationship between
height and weight.

## Two-Sample T-Test:

This test was conducted in order to determine whether the mean height of
males and females are the same.  
\
**Findings:**

|                        | Test Statistic | Degrees of Freedom | P-Value              | 95% Confidence Interval |
|------------------------|----------------|--------------------|----------------------|-------------------------|
| Linear Regression Test | 39.8477942     | 998                | 1.5319544 x 10\^-208 | (9.6501189, 10.6498089) |

Therefore, it is determined that the mean height of males and females
are not the same.

## Test for Independance:

This test was conducted in order to determine whether there is any
association between gender and amount of physical activity.  
\
**Findings:**

|                        | Test Statistic | Degrees of Freedom | P-Value   |
|------------------------|----------------|--------------------|-----------|
| Linear Regression Test | 3.226111       | 3                  | 0.1992778 |

Therefore, it is determined that gender and amount of physical activity
are independant of each other.

# Introduction:

This report aims to use multiple statistical tests to examine a data set
about the weights, heights and amount of physical activity for males and
females. There are three tests that will be conducted: linear
regression, two-sample t-test and the test for independance
($\chi^{2}$). Through conducting these tests, the report is able to
highlight the practicality and convenience of statistical tests.\
\

This report will answer three research questions, all pertaining to a
different statistical test:\
\
**Q1:** Is there a linear relationship between height and weight? This
will be answered through a linear regresssion test.\
**Q2:** Is the mean height of male and female the same? This will be
answered through a two-sample t-test.\
**Q3:** Is there any association between gender and the amount of
physical activity? This will be answered through a test for independance
($\chi^{2}$).


# Methods:

All three tests require the following structure:\
**1.** State the null and alternate hypotheses\
**2.** State the assumptions and prove them\
**3.** Calculate the observed value of the test-statistic\
**4.** Calculate the P-Value\
**5.** Calculate the confidence interval\
**6.** Conclusion to the test

\newpage

# Linear Regression Test:

## Hypothesis:

We are testing:

```{=tex}
\begin{center}
$H_{0}$ : $\beta$ = 0  against  $H_{1}$ :  $\beta$ $\neq$ 0
\end{center}
```
Where:\
$H_{0}$ is the null hypothesis and $H_{1}$ is the alternate hypothesis,\
$\beta$ is the true slope parameter as in the model Y = $\alpha$ +
$\beta$$X$ + $\epsilon$\

## Assumptions:

The assumptions for the linear regresson test are:\
1. Linearity of the data\
2. Residual errors have a mean of 0\
3. Residual errors have constant variance \
[@ASSU1]\
\
**Checking Assumptions:**

```{r assumption_1, echo=FALSE, fig.cap="Proving Assumptions for Test 1",  out.width="75%"}
residuals1 <- lm(formula = height ~ weight, data = a)$residuals #this creates a simple regression model
par(mfrow = c(2,2))
    plot(x = a, main = "height vs weight") #this plots  height vs weight
    plot(x = residuals1, main = "Regression Residuals vs Fitted Values", xlab = "Fitted Values", ylab = "Regression Residuals") #this plots the scatter plot
    hist(x = residuals1, main = "Histogram of the Regression Residuals", xlab = "Residuals") #this plots the histogram
```

It is clear from Figure 1 that all three assumptions have been proven,
therefore the test can be conducted.

## Observed Value of Test Statistic:

```{r teststat1, echo=FALSE}
lmtest <- lm(formula = height ~ weight, data = a)
tvalue <- summary(lmtest)$coefficients[6]
df <- lmtest$df.residual #this calculates the degrees of freedom
```

The observed value of the test statistic for the linear regression test
is:

$$\tau_{obs} = \frac{\hat{\beta}}{s_{Y \mid X}\sqrt{S_{XX}}}\sim{t_{df}}$$
Where, $\beta$ is the true slope parameter, jbhbcehb\
\

$\therefore$ $\tau_{obs}$ = `r tvalue`\
under $\sim$ $H_0$ and when df = `r df`

## Calculation of P-Value:

```{r p-value1, echo=FALSE}
  pvalue1 <- summary(lmtest)$coefficients[8] #this calculates the p-value
```

The P-value for the linear regression test is:\
\
P-value = $P(|t_{`r df`}| \geq `r tvalue`) = `r pvalue1`$

## Confidence Intervals:

```{r ci_1, echo=FALSE}
    ci1 <- confint.lm(lm(formula = height ~ weight, data = a), level = 0.95)[c(2,4)] ##this calculates the confident interval
```

The 95% confidence intervals for the linear regression test are:\
\
(`r ci1`)

## Conclusion:

```{r conc1, echo=FALSE}
conc1 <- matrix(c(tvalue, df, pvalue1, ci1), ncol=5,byrow=TRUE)
colnames(conc1) <- c("Test Statistic", "Degrees of Freedom", "P-Value", "Lower CI", "Upper CI")
rownames(conc1) <- c("Linear Regression Test")
conc1 <- as.table(conc1)
kable(conc1, caption = "Results for Test 1")
```

*Note:* The exact value of the P-Value is 4.8071712 x $10^{-59}$  
\
It can be concluded that as the P-value is less than 0.05, there is
enough evidence to reject the null hypothesis. Therefore, there **is** a
linear relationship between height and weight. Furthermore, the 95%
confidence interval is (0.2648081, 0.3324183), suggesting that 95 out of
100 times, the estimate will fall between the upper and lower value
[@CONC1]

\newpage

# Two-Sample T-Test:

## Hypothesis

We are testing:

```{=tex}
\begin{center}
$H_{0}$ : $\mu1 = \mu2 $  against  $H_{1}$ :  $\mu_1  \neq \mu_2 $
\end{center}
```
Where:\
$H_{0}$ is the null hypothesis and $H_{1}$ is the alternate hypothesis,\
$\mu1$ is the mean height of male and $\mu2$ is the mean height of
female

## Assumptions:

The assumptions for the two-sample t-test are:\
1. Data values must be independent\
2. Data in each group are normally distributed\
3. The variances for the two independent groups are **equal** [@ASSU2]\
\

**Checking Assumptions:**

```{r assumption_2a, include=FALSE}
  t2_p1 <- ggqqplot(b, x = "height", facet.by = "gender" ) + ggtitle("Proving Normality of the Data")
  t2_p2 <- ggplot(b, aes(x = gender, y = height)) +
    geom_boxplot() + #this plots the boxplot to prove equal variances
    ggtitle("Proving Equal Variance") #this is the title of the plot
   t2_p1
   t2_p2
```

```{r assumption_2b, echo=FALSE, fig.show="hold", fig.cap="Proving Assumptions for Test 2", out.width="50%"}
  t2_p1
  t2_p2
```

It is clear from Figure 2 that all three assumptions have been proven,
therefore the test can be conducted.

\newpage

```{r teststat2, echo=FALSE}
 ttest <- t.test(test2_male$height, test2_female$height, var.equal = TRUE)
  t_tvalue <- ttest$statistic
  t_df <- ttest$parameter
```

## Observed Value of Test Statistic:

The observed value of the test statistic for the two-sample t-test is:\
$$\tau_{obs} = \frac{\overline{X}_1-\overline{X}_2}{S_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}}\sim{t_{{n}_{1}+{n}_{2}-2}}$$
Where $\bar{X}_1$ and $\bar{X}_2$ are the sample means, $n_1$ and $n_2$
are the sample sizes, and where $S_p^{2}$ (observed pool variance) is
calculated by: $$
S_p^2 = \frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2} 
$$ $\therefore$ $\tau_{obs}$ = `r t_tvalue`\
under $\sim$ $H_0$ and when df = `r t_df`

## Calculation of P-Value:

```{r p-value2, echo=FALSE}
  t_pvalue <- ttest$p.value
```

The P-value for the two-sample t-test is:\
\
P-value = $P(|t_{`r t_df`} | \geq `r t_tvalue`) = `r t_pvalue`$

## Confidence Intervals:

```{r ci_2, echo=FALSE}
  t_ci <- ttest$conf.int
```

The 95% confidence intervals for the two-sample t-test are:\
\
(`r t_ci`)

## Conclusion:

```{r conc2, echo=FALSE}
conc2 <- matrix(c(t_tvalue, t_df, t_pvalue, t_ci), ncol=5,byrow=TRUE)
colnames(conc2) <- c("Test Statistic", "Degrees of Freedom", "P-Value", "Lower CI", "Upper CI")
rownames(conc2) <- c("Two-Sample T-Test")
conc2 <- as.table(conc2)
kable(conc2, caption = "Results for Test 2")
```

*Note:* The exact value of the P-Value is 1.5319544 x $10^{-208}$  
\

It can be concluded that as the P-value is less than 0.05, there is
enough evidence to reject the null hypothesis. Therefore, the mean
height of male and female are not equal. Furthermore, the 95% confidence
interval is (9.6501189, 10.6498089), suggesting that 95 out of 100
times, the estimate will fall between the upper and lower value [@CONC1]

\newpage

# Test for Independance:

## Hypothesis:

We are testing:

$H_{0}$ : Gender and amount of physical activity are independant of each
other *against*\
$H_{1}$ : Gender and amount of physical activity are depdnant on each
other

Where:\
$H_{0}$ is the null hypothesis and $H_{1}$ is the alternate hypothesis

## Assumptions:

The assumptions for the test for independance are:\
1. Study groups must be independent\
2. No cell shall have an expected value of less than 1\
3. At least 80% of the cells must have an expected value greater than 5 
[@ASSU3]

**Checking Assumptions:**

```{r assumption_3a, echo=FALSE}
kable(gender_phys, caption = "Observed Values (Contingency Table)")
```

```{r assumptions_3b, echo=FALSE}
none_total <- gender_phys[1,1] + gender_phys[1,2]
moderate_total <- gender_phys[2,1] + gender_phys[2,2]
intense_total <- gender_phys[3,1] + gender_phys[3,2]
female_total <- gender_phys[1,1] + gender_phys[2,1] + gender_phys[3,1]
male_total <- gender_phys[1,2] + gender_phys[2,2] + gender_phys[3,2]
total_total <- female_total + male_total

r1_col1 <- (none_total*female_total)/(total_total)
r2_col1 <- (moderate_total*female_total)/(total_total)
r3_col1 <- (intense_total*female_total)/(total_total)
r1_col2 <- (none_total*male_total)/(total_total)
r2_col2 <- (moderate_total*male_total)/(total_total)
r3_col2 <- (intense_total*male_total)/(total_total)

expected_values <- matrix(c(r1_col1, r1_col2, r2_col1, r2_col2, r3_col1, r3_col2), ncol=2,byrow=TRUE)
colnames(expected_values) <- c("Female", "Male")
rownames(expected_values) <- c("None","Moderate","Intense")
expected_values <- as.table(expected_values)
kable(expected_values, caption = "Expected Values")

```

It is clear from Table 7 that assumption #2 and #3 have been proven. It
is also known that assumption #1 is true, therefore the test can be
conducted.

\newpage

## Observed Value of Test Statistic:

```{r teststat3, echo=FALSE}
 chitest <- chisq.test(gender_phys)
 xvalue <- chitest$statistic
 chi_df <- chitest$parameter
```

The observed value of the test statistic for the test for for
independance is:\
$$\tau_{obs} = \sum_{i=1}^{r} \sum_{j=1}^{c}\frac{(O_{ij}-E_{ij})^2}{E_{ij}}\sim{\chi^2_{(r-1)(c-1)}}$$
Where, $O_i$ is the observed value, $E_i$ is the expected value,
r=number of rows, c=number of columns\
$\therefore$ $\tau_{obs}$ = `r xvalue`\
under $\sim$ $H_0$ and when df = `r chi_df`

## Calculation of P-Value:

```{r p-value3, echo=FALSE}
  chi_pvalue <- chitest$p.value
```

The P-value for the test for independance is:\
\
P-value = $P(|t_{`r chi_df`} | \geq `r xvalue`) = `r chi_pvalue`$

## Conclusion:

```{r conc3, echo=FALSE}
conc3 <- matrix(c(xvalue, chi_df, chi_pvalue), ncol=3,byrow=TRUE)
colnames(conc3) <- c("Test Statistic", "Degrees of Freedom", "P-Value")
rownames(conc3) <- c("Test for Independance")
conc3 <- as.table(conc3)
kable(conc3, caption = "Results for Test 3")
```

It can be concluded that as the P-value is greater than 0.05, there is
not enough evidence to reject the null hypothesis. Therefore, gender and
amount of physical activity are independant of each other.

\newpage

# Conclusion

The aim of this report was the answer the following research questions:  
\
**Q1:** Is there a linear relationship between height and weight? (By
conducting a linear regresssion test).\
**Q2:** Is the mean height of male and female the same? (By conducting a
two-sample t-test).\
**Q3:** Is there any association between gender and the amount of
physical activity? (By conducting a test for independance ($\chi^{2}$)).

After stating the hypotheses, proving the assumptions and performing the
statistical analysis for all three of the tests, it can be concluded
that:

| **Research Question** | **Statistical Test**   | **Answer to Research Question**                                      |
|----------------|----------------|----------------------------------------|
| #1                    | Linear Regression Test | There is a linear relationship between height and weight             |
| #2                    | Two-Sample T-Test      | The mean height of males and females are not the same                |
| #3                    | Test for Independance  | Gender and amount of physical activity are independant of each other |

Therefore, through a real life scenario, this report was able to highlight the practicality and convenience of statistical tests.

\newpage

# References
